---
title: "Effect commands"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Effect commands}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What are effect commands

A pattern table is composed of cells telling the tracker which notes to play
in which order. It also has a field for effect commands. It is often represented
as a three digit hexadecimal value. The first hex digit represents the effect
command, the latter represent parameter values used by the command. The
table below summarises all available effects in ProTracker and is followed
by sections describing them in more detail and providing examples.

```{r tab-effect-commands, echo=FALSE, message=FALSE, warning=FALSE}
library(ProTrackR2)
library(kableExtra)
data(effect_commands)
effect_commands |> kbl()
```

## Extracting and updating effect commands

To get an effect command from a ProTracker module, you can simply call
`pt2_command()` to a `pt2cell` or `pt2celllist` class object (see `vignette("s3class")` for
a ProTrackR2 object tree). There, are several ways of extracting cells as
documented in `vignette("sel_assign")`. In the example below there is one way shown
of getting the effect commands of the first 8 rows of the second column in
the third pattern.

```{r select-command}
mod <- pt2_read_mod(pt2_demo())
pt2_command(
  mod$patterns[[ 3 ]] [ 1:8, 2 ])
```

You can also assign new effect commands to your selection like this:

```{r assign-command}
pt2_command(
  mod$patterns[[ 3 ]] [ 1:8, 2 ]) <- "C10"
```

<!-- TODO combination with instrument number -->
<!-- TODO include details and examples -->