---
title: "ProTrackR vs ProTrackR2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ProTrackR vs ProTrackR2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## ProTracker2 and its predecessor

The predecessor to this package ([ProTrackR](https://pepijn-devries.github.io/ProTrackR/)) was
entirely programmed in R. Although technically possible, it was challenging and slow with recursive algorithms.
The new version is a complete overhaul in C/C++, based on Olav S&oslash;rensen's
[ProTracker clone](https://16-bits.org/pt2.php). With it, came some design changes which are worth mentioning
here.

## Overview of differences

The table below summarises the differences between ProTrackR2 and its predecessor.

```{r tab-dif, echo=FALSE, message=FALSE, warning=FALSE}
library(kableExtra)
load("../tests/testthat/_openmpt/testcasestate.Rdata")
testresult <- table(openmpt_testfiles$state)
data.frame(
  Feature    = c("[Effect commands](#effect-commands)", "[Infrastructure](#infrastructure)",
                 "[File readers](#file-readers)", "[Audio output](#audio-output)", "OpenMPT test cases"),
  ProTrackR  = c("Limited set implemented", "R script", "Optimized for format preservation",
                 "tuneR S4 Wave class", "Passes 6 out of 12 selected tests"),
  ProTrackR2 = c("All PT2.3d effects implemented", "Compiled C/C++", "Optimized for PT2.3d compatibility",
                 "audio S3 audioSample class",
                 sprintf("Passes at least %i of the %i tests", testresult[["Success"]], sum(testresult)))
) |>
  kbl()
```

## Effect commands

ProTracker uses specific codes to apply certain effects or position jumps. `ProTrackR` implements only
a subset of these effects, whereas `ProTrackR2` has implemented all ProTracker compatible effects.

The predecessor only partly implemented arpeggio and setting finetune. It did not implement
glissando, sample filtering (E8) and loop reversal (EF). All these effects are implemented in
the current package.

For a full overview of effect commands see `vignette("effect_commands")`.

## Infrastructure

By switching to `C` and `C++` compiled code, the new package gained a significant
performance boost (see [benchmark results](#benchmark)). Where the in the predecessor
the module was represented by a vector of `raw` data, an `externalptr` to a `C struct`
is used in the current package. This required a slightly different approach to handle
these objects. In order to avoid confusion about the syntax, it was completely redesigned
in the successor.

<!--
TODO
S3 class instead of S4 etc.
changes in syntax, some examples side by side
-->

### Benchmark

A benchmark test where the same module (the one provided with this package) is rendered with
both `ProTrackR` and `ProTrackR2`. The settings for both tests were similar and performed
on the same system and repeated 10 times. On average `ProTrackR2` renders 8.8 times faster
than `ProTrackR`.

## File readers

While reading ProTracker modules, the predecessor preserved the data in the file. It only
modified / fixed data when requested by the user. The current package will always sanitise
data while reading it, making it compatible with ProTracker 2.3d. The current reader is
also a bit more flexible and allows to read more exotic formats. It even allows you to
read files compressed with PowerPacker.

### Alternatives

If you want even more flexibility, check out the [openmpt package](https://pepijn-devries.github.io/openmpt/).
It uses [libopenmpt](https://lib.openmpt.org/) to read and play modules. This library
has a more extensive set of supported [file formats](https://wiki.openmpt.org/Manual:_Module_formats).
The downside is that it does not allow you to modify or save modules.

## Audio output

The predecessor used [tuneR](https://cran.r-project.org/package=tuneR) objects to store
rendered audio. In the current package we use [audio](https://cran.r-project.org/package=tuneR)
objects. This switch was made as the `S3` class objects from 'audio' are easier to handle than the
stricter and formal `S4` class objects from 'tuneR'. If you wish to use the advanced features
from the 'tuneR' package, this is still possible as both formats can be converted relatively easy.

```{r audio-tuner, eval=FALSE}
## Load demo module
mod <- pt2_read_mod(pt2_demo())

## render 'audioSample' object
mod_audio <- pt2_render(mod)

## Convert from 'audio::audioSample' to 'tuneR::Wave' object:
if (requireNamespace("tuneR")) {
  mod_tuneR <-
    tuneR::Wave(
      left      = as.integer(2^15*unclass(mod_audio[1,])),
      right     = as.integer(2^15*unclass(mod_audio[2,])),
      samp.rate = attributes(mod_audio)$rate,
      bit       = attributes(mod_audio)$bits
    )
}
```
