---
title: "ProTrackR2 S3 class objects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ProTrackR2 S3 class objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{js, echo = FALSE}
function getpatclass(i) {
  var classname= 'patcolumn' + i;
  var els = document.getElementsByClassName(classname);
  return els;
}

function highlightcol(colindex, mouseover) {
  for (j = 1; j < 5; j++) {
      var els = getpatclass(j);
      for (i = 0; i < els.length; i++) {
        if (mouseover && colindex == j) {
          els[i].style.borderLeft  = '2px solid red';
          els[i].style.borderRight = '2px solid red';
        } else {
          els[i].style.borderLeft = '0px none';
          els[i].style.borderRight = '0px none';
        }
      }
  }
}

function getpatclass2(i) {
  var classname= 'patrow' + i;
  var els = document.getElementsByClassName(classname);
  return els;
}

function highlightrow(rowindex, mouseover) {
  for (j = 0; j < 7; j++) {
      var els = getpatclass2(j);
      for (i = 0; i < els.length; i++) {
        if (mouseover && rowindex == j) {
          els[i].style.borderTop  = '2px solid orange';
          els[i].style.borderBottom = '2px solid orange';
        } else {
          els[i].style.borderTop = '0px none';
          els[i].style.borderBottom = '0px none';
        }
      }
  }
}

```

```{css, echo=FALSE}
.pt2note {
  background-color: #0000FF22;
  display:inline;
  float:left;
}

.pt2instr {
  background-color: #00FFFF22;
  display:inline;
  float:left;
}

.pt2effect {
  background-color: #00FF0022;
  display:inline;
  float:left;
}

g.grvzclick:hover {
  outline: solid 3px blue;
}

table.tab_patt td {
  border-left: 0px none;
  border-right: 0px none;
}

table.tab_patt {
  font-family: monospace;
}
```
## Outline of ProTrackR2 objects

The ProTrackR2 package has defined a number of hierarchical classes to represent ProTracker modules.
This vignette explains how these objects are organised, and how they can be used to work with
ProTracker modules in R. We start with the diagram below, which shows an object tree. It shows
how object classes are organised and related. It is similar (but not identical) to how ProTracker
module files are organised. We follow the tree to explain each object by starting at its root:
the ProTracker module.

```{r, echo=FALSE, results='asis'}
htmltools::includeHTML("../man/figures/object-tree.svg")
```

## ProTracker Module objects (`pt2mod`)

The `pt2mod` class object is a memory representation of ProTracker module files. The type of this
object is `externalptr` and points to the location in memory where the data is stored for manipulation
and/or play back.

An empty module can be created by calling `pt2_new_mod("song title")`. It can also be initiated by
reading a module from a file like this:

```{r setup}
library(ProTrackR2)
mod <- pt2_read_mod(pt2_demo())
mod
```

As you might know, external pointers cannot be accessed directly in `R`. Instead you need to use
the functions and methods presented in this package to approach it. If you want a more 'physical'
manifestation of the object, you can call `as.raw()` (which returns the raw file representation of
the object). Or write it to a file using `pt2_write_mod()`. If you want to know what the module
sounds like you can simply play it:

```{r play, eval=FALSE}
play(mod)
```

## Sample list objects (`pt2samplist`)

The sample list is a collection of distinct instruments (samples) and is represented by the `pt2samplist`
class. Essentially, it is a list of samples (`pt2samp` class objects; see below). The sample list of
a module can be obtained using the `$`-operator as shown below.

```{r sample-list}
mod$samples
```

See also `vignette("sel_assign")` for more details about selecting or replacing a sample list.
The sample list returned here is the complete list of samples, including empty sample slots.
If you want to count the number of non-empty samples in a module use `pt2_n_sample(mod)`.

### Sample objects (`pt2samp`)

ProTracker uses pulse-code modulation (PCM) mono audio samples with 8 bit depth. In this package it's
represented by the `pt2samp` class. It is a vector of `raw` values, where each value is an 8 bit PCM
sample. The `raw` form
will also hold an attribute named `"sample_info"` used to interpret the sample. The attribute
`"sample_info"` is a named list containing the following elements:

  * `length`: should be identical to the length of the vector of `raw` values.
  * `loopStart`: start position (in number of samples) of a loop in the sample (0 when loop is off).
  * `loopLength`: length (in number of samples) of a loop (2 when loop is off).
  * `fineTune`: an `integer` value (between -8 and 7) to tweak a sample when it is out of tune.
  * `volume`: volume level of the sample. An `integer` value between 0 (muted) and 64 (max).
  * `text`: 22 UTF-8 characters. Usually describing the sample or the module. It can
    also contain a message to the audience.

A sample can be initiated by reading it from an audio file using `pt2_read_sample()`. It can
also be selected from a sample list as shown below.

```{r sample}
my_sample <- mod$samples[[1]]
```

See also `vignette("sel_assign")` for more details about selecting or replacing a sample.

## Pattern list objects (`pt2patlist`)

The pattern list is a collection of distinct pattern tables and is represented by the `pt2patlist`
class. Essentially, it is a list of pattern tables (`pt2pat` class objects; see below). Note that
this list of pattern tables is not necessarily the order in which they are played. This is determined
by the pattern sequence (`pt2_pattern_table(mod)`).

The pattern list of a module can be obtained using the `$`-operator as shown below.

```{r pattern-list}
mod$patterns
```

The loaded module consists of `r length(mod$patterns)` unique pattern tables.
See also `vignette("sel_assign")` for more details about selecting or replacing a pattern list.

### Pattern objects (`pt2pat`)

A pattern table defines the rhythm and melody of a ProTracker module. The columns in a pattern
table represent the four audio channels (red borders in <a href="#tab-pat-example">Table 1</a>).
Each row (orange borders in <a href="#tab-pat-example">Table 1</a>) is subsequently played and
specifies which note (blue fields in <a href="#tab-pat-example">Table 1</a>) using a specific instrument
(sample, cyan fields in <a href="#tab-pat-example">Table 1</a>) is heard. It can also specify specific
effects such as tremolo, porta or vibrato (green fields in <a href="#tab-pat-example">Table 1</a>).
See for more details `vignette("effect_commands")`.

```{r tab-pattern, echo=FALSE, message=FALSE, warning=FALSE}
library(ProTrackR2)
mod <- pt2_read_mod(pt2_demo())

mat <-
  mod$patterns[[1]] |>
  as.character() |>
  apply(1, \(x) gsub("-", "&minus;", x))
mat <- mat[,c(1:5, 64)] |>
  apply(2, \(x) {
    strsplit(x, " ") |>
      lapply(\(y)
             sprintf("<div class='pt2note'>%s</div><div class='pt2instr'>&nbsp;&nbsp;%s&nbsp;&nbsp;</div><div class='pt2effect'>%s</div>",
                     y[1], y[2], y[3])) |> unlist()
  })
mat <- rbind(c(0:4, 63), mat)
mat[,5] <- "..."

mat <- outer(1:5, 1:6, function(X, Y) {
  sprintf("<td class='patcolumn%i'> %s </td>", X - 1, mat[cbind(X, Y)])
}) |>
  apply(2, paste0, collapse = "\n") |>
  sprintf(
    fmt = sprintf("<tr class = 'patrow%i'%s%%%%s> %%s </tr>",
                  1:6,
                  c("", "", "", "", " style='text-align: center;'", ""))
) |>
  paste0(collapse = "\n")

mat <-
  do.call(sprintf, c(
    list(fmt = mat),
    sprintf(" onmouseleave='highlightrow(%i, false);' onmouseenter='highlightrow(%i, true);'",
            1:6, 1:6)))

hd <-
  c("<th> Row </th>",
    sprintf("<th onmouseleave='highlightcol(%i, false);' onmouseenter='highlightcol(%i, true);'> Channel %i </th>", 1:4, 1:4, 1:4)) |>
  paste0(collapse = "\n") |>
  sprintf(fmt = "<thead>\n<tr>\n%s</tr></thead>")
cp <- "<caption><span id='tab-pat-example'>Table 1: </span>An example of a pattern table. Note that rows 4 to 62 are omitted.</caption>"
sprintf("<table class='tab_patt'>%s\n\n%s\n<tbody>\n%s\n</tbody>\n</table>",
        cp, hd, mat) |>
  htmltools::HTML()
```

A ProTracker pattern is represented by the `pt2pat` class. It is a `vector` of `raw` values
that either represents the file structure of a pattern or its representation in memory used by
the play back routine.

A pattern can be initiated with `pt2_new_pattern()`. This produces a `raw` representation
of a pattern. Here the `logical` argument `compact` determines if the notation is compact (`TRUE`)
as is the case in a file, or not (`FALSE`) as in memory. This is stored as an attribute to
the object for its correct interpretation.

It can also be selected from an existing module as shown below.

```{r pattern}
my_ptn <- mod$patterns[[1]]
```

Consult `vignette("sel_assign")` for more information about selecting and replacing patterns.

### Cell list objects (`pt2celllist`)

The cell list is a collection of cells from a pattern table and is represented by the `pt2celllist`
class. Essentially, it is a list of cells (`pt2cell` class objects; see below).
The cell list can be obtained using the `[`-operator from a pattern as shown below.

```{r celllist}
## This selects all cells in the first two channels
## of a pattern
my_cells <- my_ptn[,1:2]
```

Consult `vignette("sel_assign")` for more information about selecting and replacing cell lists.

### Cell objects (`pt2cell`)

A cell is an elementary unit in a pattern table and contains information about which note to play
(blue fields in <a href="#tab-pat-example">Table 1</a>),
which instrument (sample cyan fields in <a href="#tab-pat-example">Table 1</a>)
and what kind of effect to apply (green fields in <a href="#tab-pat-example">Table 1</a>).
It is represented by the `pt2cell` class
and is represented by a vector of
`raw` values as stored in file (compact) or in memory (not compact). A cell can be selected from
a cell list as shown below.

```{r cell}
my_cells[[1]]
```

See `vignette("sel_assign")` for more information about selecting and replacing cell values.

### Command objects (`pt2command`)

The command objects are used to represent ProTracker effect commands used in specific cells
(see `vignette("effect_commands")`). It is essentially either a `pt2cell` or a `pt2celllist`
object with `pt2command` as a child class.

It can be used to select or replace effect commands (see `pt2_command()` or `vignette("sel_assign")`).
From the diagram at the top of this document you can see that cells also contain instrument and
note information. These don't inherit specific classes as they are represented by simple
`integer` and `character` values. For more information about those please check `pt2_instrument()` and
`pt2_note()`.
